Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-19 Manuscript/manuscript.tex
Add references for the cationic surfactant simulations
Some updates in binding affinity details and stoichiometry sections. I will continue now with the residence times.

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
\todo{JOE: We cannot present an intricate cool new model for a lipid and claim at the same time that we can't simulate bromide. }
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

@@

\subsection{Cation binding affinities to POPC membrane validated through lipid electrometer}

@@

\subsection{Cation binding to POPC membrane in atomistic detail}

@@

The binding affinities of Ca$^{2+}$ ions to POPC bilayer in different simulation 
models were quantified by calculating the relative surface excess using Eq.~\ref{surfexcess} 
and the bulk concentrations determined from the density profiles in Fig.~\ref{fig:cacl-dens}. 
All these simulation have the same, $350\,\mathrm{mM}$, concentration of CaCl$_2$ with 
respect to water, i.e., the concentration before solvating the lipids in experiments.
As expected from the changes of the lipid headgroup order parameters in Fig. \ref{fig:delta_ordPar_CaCl}, 
the relative surface excess for the ECC-lipid model, $\Gamma_{\rm Ca}^w = 0.07 \pm 0.01 \rm{nm}^{-2}$,
is significantly smaller than for the Lipid14 model with \AA{}qvist ions, $\Gamma_{\rm Ca}^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_{\rm Ca}^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table SAMULI: They are now in the table \ref{tab:binding} 
together with probabilities of the calcium binding probabilities to different POPC oxygens. 
However, I am not sure if the table is useful in the end.}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) 
calculated using density profiles in Fig.~\ref{fig:nacl-dens}
is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which predict a specific binding of sodium on 
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
%
%This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
%which show the density profile of ions and water along the membrane normal. 
%
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion above to be finished after this.}

@@

The probabilities for a bound \ce{Ca^{2+}} ion to interact with
different oxygen moieties of POPC 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by counting the contacts within the distance of $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}. 
%
%SAMULI: I commented out the average number of contacts. All the essential information
%for this is in the probabilities.
%
%In contrast to sodium, the dominant contribution
%to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
%% Numbers from simulation with 346mM conc. CaCl2
%The average number of contacts per lipid 
%between \ce{Ca^{2+}} at $290\,\mathrm{mM}$ concentration and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are considered 
%%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
%
% Numbers from simulation with 980/845mM conc. CaCl2
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
The results in Table \ref{tab:binding} show that almost all (99\%) of the 
bound Ca$^{2+}$ ions are in contact with the phosphate oxygens and a quarter
of them (25\%) interacts with the carbonyl oxygens. Thus, the interactions between
Calcium ions with only carbonyl oxygens are very unlikely (1\%), but significant
amount of the ions interact with both, phosphate and carbonyl oxygens.
The individual lipids or acyl chains were not identified in the analysis,
thus the ion contacts with both oxygen moieties may be inter or intra molecular
and occur with the carbonyls in the either of the acyl chains.
The most likely contacts between Ca$^{2+}$ ions and phosphate oxygens is visualized with
the probability isodensity contours in Fig.~\ref{fig:volmaps}.
While the higher concentrations of \ce{CaCl2} naturally increased the amount of contacts per lipid,
%for $850\,\mathrm{mM}$ concentration of \ce{CaCl2}
%the average number of contacts per lipid is increased to
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $, 
the distribution of contacts between phosphate and carbonyl oxygens was not affected.
%and density profiles of ions and water along the membrane normal in Fig.~\ref{fig:cacl-dens}.

@@

Even thought \ce{Na+} ions bind to POPC bilayer with a very low probability,
it was possible to analyze the distribution of contacts from simulations with a large
sodium concetration of~1\,M.
%the average number of contacts per lipid with any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%this number decreases to 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are counted 
%it is even lower, 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
The results show that the sodium ions bind purely to the phosphate oxygens 
of POPC with the probability of 55\% and purely to the carbonyl oxygens
with the probability of 20\%, leaving the probability of 25\% for the
simultaneous binding to both, phosphate and carbonyl oxygens.  
%This is in line with the calculated negative value of 
%relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
%their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
%suggesting that the populations of bound sodium cations 
%naturally decrease with the decreasing density of water.

@@

\subsection{Stoichiometry of binding of \ce{Na+} and \ce{Ca^{2+}} cations to POPC membrane}

@@

The same experimental data for the lipid headgroup order parameters
used in this work to validate the simulation models in Fig. \ref{fig:delta_ordPar_CaCl},
has been previous interpreted by using simple binding models to determine the
binding constants and stoichimetry of \ce{Ca^{2+}} binding to PC lipid
bilayers~\cite{altenbach84,macdonald87}. The experimental data was best
explained by the ternary complex binding model with a binding
stoichiometry of one \ce{Ca^{2+}} per two POPC lipids~\cite{altenbach84}.
However, also a Langmuir adsorption model (i.e. \ce{Ca^{2+}}:POPC stoichiometry of 1:1) 
was successfully fit to the experimental data when only low concentrations of \ce{CaCl2} were considered~\cite{macdonald87}.
The results based on these binding models and the measured headgroup order parameter changes
are also inluded in the tabulated experimental binding constants~\cite{marsh13}.

Since the experimentally measured changes of the POPC lipid headgroup order parameters
as a function of CaCl$_2$ concentration are reproduced by the ECC-lipid model,
it can be used to give a more versatile interepration of the binding stoichiometry than
the previously used simple binding models. To directly evaluate the
stoichiometry from simulations we calculated the relative propensities of \ce{Ca^{2+}}:POPC
complexes with various stoichiometries by counting the individual lipid
molecules within the distance of $0.3\,\mathrm{nm}$ from each \ce{Ca^{2+}} ion.
The results from POPC simulation with 285~mM concetration of CaCl$_2$ in
Fig.~\ref{fig:cacl_complexes} show the largest propensity (42\%) for the
ternary complex with the \ce{Ca^{2+}}:POPC stoichiometry of 1:2, but 
the probabilities for the complexes with stoichimetries of 1:1~(30\%)~and
1:3~(28\%)~are only slightly lower,
suggesting more complex binding stoichiometry than the simple ternary complex model.
%The observed propensities of
%$42\%$ for two lipids,
%$30\%$ for one lipid, and
%$28\%$ for three lipids 
\todo{Also the larger complexes were analysed, right? So we can write ''Larger complexes were not existent in simulation.'', or something like that?}
%The symmetry of the distribution of complexes,
%from simulation as shown in Figure~\ref{fig:cacl_complexes} --   
%i.e.,
However, one Calcium binds roughly to two lipids on average also in simulations,
because the probabilities of forming the complexes with 1 or 3 lipids are almost equal.  
This probably explains the successfull fit of the ternary complex model to the
experimental data. 
\todo{I think that we either have to do the analysis in SI properly,
  i.e., using the real $C_I$ from simulations or remove the sentence below and
the related concent from SI.}
Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/ipython_nb/stoichiometry_NaCl-CaCl2_comparison_Ecc-lipids.pdf} \\
  \caption{\label{fig:cacl_complexes}
      Relative probabilities of existence of \ce{Na^{+}} or \ce{Ca^{2+}} complexes
      with a certain number of POPC lipids. 
      \ce{Na^{+}} complexes were evaluated from the simulation with 1~M concentration;
      and \ce{Ca^{2+}} complexes were evaluated from the simulation with 287~mM concentration.
  }
\end{figure}

>>>>

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters together with the simulation data are available
at~\cite{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
\todo{JOE: We cannot present an intricate cool new model for a lipid and claim at the same time that we can't simulate bromide. }
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

@@

\subsection{Binding affinities to POPC membrane validated through lipid electrometer}

@@

\subsection{Affinity and binding cites to POPC membrane}

@@

The binding affinities of Ca$^{2+}$ ions to POPC bilayer in different simulation 
models were quantified by calculating the relative surface excess from Eq.~\ref{surfexcess} 
using the bulk concentrations determined from the density profiles in Fig.~\ref{fig:cacl-dens}. 
The relative surface excess values, shown in Table~\ref{tab:binding},
were calculated from simulations with the same, $350\,\mathrm{mM}$, concentration of CaCl$_2$ with 
respect to water, i.e., the concentration before solvating the lipids in experiments.
As expected from the changes of the lipid headgroup order parameters in Fig. \ref{fig:delta_ordPar_CaCl}, 
the relative surface excess for the ECC-lipid model, $\Gamma_{\rm Ca}^w = 0.07 \pm 0.01 \rm{nm}^{-2}$,
is significantly smaller than for the Lipid14 model with \AA{}qvist ions, $\Gamma_{\rm Ca}^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_{\rm Ca}^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table SAMULI: They are now in the table \ref{tab:binding} 
together with probabilities of the calcium binding probabilities to different POPC oxygens. 
However, I am not sure if the table is useful in the end.}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) 
calculated using density profiles in Fig.~\ref{fig:nacl-dens}
is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which predict a specific binding of sodium on 
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
%
%This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
%which show the density profile of ions and water along the membrane normal. 
%
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion above to be finished after this.}

@@

The probabilities for a bound \ce{Ca^{2+}} ion to interact with
different oxygen moieties of POPC 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by counting the contacts within the distance of $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}. 
%
%SAMULI: I commented out the average number of contacts. All the essential information
%for this is in the probabilities.
%
%In contrast to sodium, the dominant contribution
%to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
%% Numbers from simulation with 346mM conc. CaCl2
%The average number of contacts per lipid 
%between \ce{Ca^{2+}} at $290\,\mathrm{mM}$ concentration and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are considered 
%%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
%
% Numbers from simulation with 980/845mM conc. CaCl2
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
The results in Table \ref{tab:binding} show that almost all (99\%) of the 
bound Ca$^{2+}$ ions are in contact with the phosphate oxygens and a quarter
of them (25\%) interacts with the carbonyl oxygens. Thus, the interactions between
Calcium ions with only carbonyl oxygens are very unlikely (1\%), but significant
amount of the ions interact with both, phosphate and carbonyl oxygens.
The individual lipids or acyl chains were not identified in the analysis,
thus the simultaneous ion contacts with both oxygen moieties may be inter or intra molecular
and occur with the carbonyls in the either of the acyl chains.
The most likely contacts between Ca$^{2+}$ ions and phosphate oxygens is visualized with
the probability isodensity contours in Fig.~\ref{fig:volmaps}.
While the higher concentrations of \ce{CaCl2} naturally increased the amount of contacts per lipid,
%for $850\,\mathrm{mM}$ concentration of \ce{CaCl2}
%the average number of contacts per lipid is increased to
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $, 
the distribution of contacts between phosphate and carbonyl oxygens was not affected.
%and density profiles of ions and water along the membrane normal in Fig.~\ref{fig:cacl-dens}.

@@

Even thought \ce{Na+} ions bind to POPC bilayer with a very low probability,
it was possible to analyze the distribution of contacts from simulations with a large
sodium concetration of~1\,M.
%the average number of contacts per lipid with any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%this number decreases to 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are counted 
%it is even lower, 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
The results show that \ce{Na+} ions bind only to the phosphate oxygens 
of POPC with the probablity of 55\% and only to the carbonyl oxygens with the
probability of 20\%, leaving the probability of 25\% for the
simultaneous binding to both.%, phosphate and carbonyl oxygens.  
%This is in line with the calculated negative value of 
%relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
%their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
%suggesting that the populations of bound sodium cations 
%naturally decrease with the decreasing density of water.

@@

\subsection{Binding stoichiometry to POPC membrane}

@@

Simple binding models have been previous used to interpreted 
the same experimental data, which is used in this work to validate
the simulation models in Fig. \ref{fig:delta_ordPar_CaCl}~\cite{altenbach84,macdonald87}.
The experimental results from the lipid headgroup order parameters and from the
atomic absorption spectroscopy were best
explained by the ternary complex binding model with a binding
stoichiometry of one \ce{Ca^{2+}} per two POPC lipids~\cite{altenbach84}.
However, also a Langmuir adsorption model (i.e. \ce{Ca^{2+}}:POPC stoichiometry of 1:1) 
was successfully fit to the experimental data when only low concentrations of \ce{CaCl2} were considered~\cite{macdonald87}.
The results based on these binding models and the measured headgroup order parameter changes
are also inluded in the tabulated experimental binding constants~\cite{marsh13}.

Since the same experimental data is reproduced by the ECC-lipid model,
it can be used to give a more versatile interepration of the binding stoichiometry than
the previously used simple binding models~\cite{altenbach84,macdonald87}. To directly evaluate the
stoichiometry from simulations we calculated the relative propensities of \ce{Ca^{2+}}:POPC
complexes with various stoichiometries by counting the individual lipid
molecules within the distance of $0.3\,\mathrm{nm}$ from each \ce{Ca^{2+}} ion.
The results from POPC simulation with 285~mM concetration of CaCl$_2$ in
Fig.~\ref{fig:cacl_complexes} show the largest propensity (42\%) for the
ternary complex with the \ce{Ca^{2+}}:POPC stoichiometry of 1:2, but 
the probabilities for the complexes with the stoichimetries of 1:1~(30\%)~and
1:3~(28\%)~are only slightly lower,
suggesting more complex binding stoichiometry than the simple ternary complex model.
%The observed propensities of
%$42\%$ for two lipids,
%$30\%$ for one lipid, and
%$28\%$ for three lipids 
However, one Calcium binds roughly to two lipids on average also in simulations,
because the probabilities of forming the complexes with 1 or 3 lipids are almost equal.  
This probably explains the successfull fit of the ternary complex model to the
experimental data. 
Complexes with more than three lipids per one Calcium ion were not observed in 
simulations with the ECC-lipids model.
\todo{I think that we either have to do the analysis in SI properly,
  i.e., using the real $C_I$ from simulations or remove the sentence below and
the related concent from SI.}
Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
\begin{figure}[tbp]
  \centering
  \includegraphics[width=6.0cm]{../Fig/ipython_nb/stoichiometry_NaCl-CaCl2_comparison_Ecc-lipids.pdf} \\
  \caption{\label{fig:cacl_complexes}
      Relative probabilities of existence of \ce{Na^{+}} or \ce{Ca^{2+}} complexes
      with a certain number of POPC lipids. 
      \ce{Na^{+}} complexes were evaluated from the simulation with 1~M concentration;
      and \ce{Ca^{2+}} complexes were evaluated from the simulation with 287~mM concentration.
  }
\end{figure}
